AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction

### 2018-03-25 doc/ms.tex
Downloaded manuscript from overleaf

\documentclass[preprint, a4paper]{elsarticle}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{dsfont}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[left=2.2cm, right=2.2cm, bottom=3cm, top=2.5cm]{geometry}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{pdflscape}
\usepackage{gensymb} % For degree symbol
%\usepackage{lineno,hyperref}
%\modulolinenumbers[5]

\journal{}

@@

\begin{abstract}
AMORPH utilizes a new Bayesian statistical approach to interpreting X-ray diffraction results of partially amorphous materials. AMORPH fits X-ray diffraction patterns with a mixture of gaussians,
which allows us to simultaneously infer all of the model parameters and quantify their
uncertainties.
The program simulates background patterns previously manually applied, providing reproducible results, and significantly reducing inter- and intra-user biases. This approach allows for the quantification of amorphous and crystalline materials and for the characterization of the amorphous component, including parameters of centre of mass, width, skewness, and nongaussianity of the amorphous component. Results demonstrate the applicability of this program for calculating amorphous contents of volcanic materials and independently modeling their properties in compositionally variable materials.
\end{abstract}

@@

%\linenumbers

\section{Introduction}
Quantification and characterization of amorphous materials is an important area of research in
terrestrial and planetary sciences \citep[e.g.,][]{schmidt2009, dehouck2014, wall2014, morris2016, zorn2017}.
Powder X-ray diffraction (XRD), although primarily a tool for characterization of crystalline materials,
also provides a means to investigate amorphous materials \citep[e.g.,][]{rowe2012, achilles2013}. X-ray diffraction of rocks/soils containing amorphous materials (e.g. volcanic
glass, amorphous silica, and organics) produces a diffraction pattern which is a combination of the
crystalline and amorphous phases. Recent studies have debated the nature of amorphous materials in
planetary sciences \citep[e.g.][]{schmidt2009, dehouck2014}, and a growing number of
applications to terrestrial volcanics have presented an avenue of expansion for XRD methodologies
for amorphous material analysis \citep{ellis2015, kanakiya2017, zorn2017}.
Here we present a new statistical approach and computer program, AMORPH, which utilizes Bayesian inference to determine
the crystallinity, and other characteristics, of partially amorphous materials from X-ray diffraction patterns, as well as the uncertainty in these quantities.

Several techniques currently exist for the quantification of amorphous materials in heterogeneous (amorphous + crystalline) materials. However, these
each present their own limitations. The two most commonly used approaches in recent literature
either rely on a calibration curve for different crystallinities \citep[calibration method;][]{ rowe2012} or a combined Rietveld-Reference Intensity Ratio (RIR) method \citep{gualtieri1996, gualtieri2000}. While the combined Rietveld-RIR procedure allows for quantification of both the amorphous 
and crystalline materials, for proper characterization it requires a library of pure components,
spiked with a reference material (e.g. corundum). Sample spiking presents additional complications
as it reduces the observed intensity of the amorphous material thus reducing the accuracy of
pattern matching to determine abundances. An Excel based adaptation of this pattern matching
technique \citep[FULLPAT;][]{chipera2002} is routinely utilized for characterizing remotely acquired X-ray
diffraction patterns, such as those recovered from the Mars Science Laboratory (MSL) Curiosity rover \citep{bish2013}. The problem faced by RIR approaches is they are based on modeling of an observed 
diffraction pattern compared to library measurements, and therefore do not allow an independent
assessment of the characteristics of the amorphous material. In comparison, the calibration method
focuses solely on the quantification of the amorphous material, utilizing the integrated counts
associated with the amorphous and crystalline componentry.

@@

\section{Bayesian inference}\label{sec:bayes}
AMORPH is founded upon Bayesian inference
\citep{gregory2005bayesian, o2004kendall, sivia2006data}, where
probability distributions are used to model
states of uncertainty about the values of unknown quantities.
Bayesian inference is often implemented computationally with
Markov Chain Monte Carlo (MCMC) methods \citep{mackay2003}, and AMORPH is no exception.
Loosely speaking, the goal is to fit a dataset with a model by exploring
those regions of the model's parameter space that are consistent with the
data. The model assumptions are described in detail in
Section~\ref{sec:model}, and the computation is performed with
Diffusive Nested Sampling, an advanced MCMC method \citep{dns, dnest4}.

@@

The prior describes an initial state of uncertainty about the
parameter values, and is often a wide probability distribution.
The prior is supplemented with $p(\data | \params, \info)$, sometimes
called a `sampling distribution',
which describes the uncertainty about the data which we {\em would} have,
if we knew the parameters (and the prior information), as a function of
the parameters. This describes uncertainty about what data will be observed,
but encodes some knowledge about the fact that the data has something to do with the
parameters $\params$. For example, $p(\data | \params, \info)$ often encodes
the assumption that the data has noise in it, and the specific noise values
are unknown a priori.

@@

When $\params$ is a large collection of parameters, the posterior distribution
is a probability distribution over a high-dimensional parameter space. To
represent it computationally, Markov Chain Monte Carlo (MCMC) methods are
often used to generate samples from the parameter space, according to the
posterior distribution. The output is a collection of plausible values of the
parameters which can be used to approximate posterior probabilities.
For example, the probability some parameter $\phi$ is greater than 3.4
is approximately the proportion of the posterior samples that have
$\phi > 3.4$. While many data analysis procedures work by finding the
best fitting values of the parameters (i.e., a maximum likelihood estimate),
Bayesian inference via MCMC involves finding a sampling of plausible values
for the parameters.

@@

\subsection{The background component}
The background component is assumed to be piecewise linear, with constant
slope
{between $\x=\x_{\rm min}$ and $\x=10$\degree, between $\x=10$\degree~and $\x=40$\degree,
and between $\x=40$\degree~and $\x=\x_{\rm max}$}. These values are selected to maintain a linear background over the interval from 10--40\degree, the predominant range of amorphous materials analyzed using a Cu K$\alpha$ x-ray source, while still allowing for some flexibility in the structure of the background outside this range.

We parameterized the background component using
five parameters. Firstly, there is a mean level $b$, which sets the
typical amplitude of the background component. There are also four parameters
$n_1, ..., n_4$, which determine the offset of four `control points'
from the mean level. The positions of the four control points are then:
\begin{align}
\left(\x_{\rm min}, be^{n_1}\right) \nonumber\\
\left(10, be^{n_2}\right) \nonumber\\
\left(40, be^{n_3}\right) \nonumber\\
\left(\x_{\rm max}, be^{n_4}\right). \nonumber
\end{align}
The value of the background component at any other point is found by
linear interpolation of these control points.
Figure~\ref{fig:background} shows three example
background curves generated from this parameterization, with typical
values of the parameters.

@@

\subsection{The crystalline component}
We assume that there is some number $N_{\rm xtal}$ of narrow peaks,
each of which has a gaussian shape. The $i$th peak is parameterized
by its central position $X_{i, \rm xtal}$,
amplitude $A_{i, \rm xtal}$, and width $W_{i, \rm xtal}$.
The parameter vector for crystalline component is
\begin{align}
\params_{\rm xtal} &=
  \{N_{\rm xtal}, X_1, X_2, ..., X_{N, \rm xtal},
    A_1, A_2, ..., A_{N, \rm xtal},
    W_1, W_2, ..., W_{N, \rm xtal}\}
\end{align}
and the total shape contributed by the
crystalline component is a sum of the $N_{\rm xtal}$ gaussians:
\begin{align}
f_{\rm xtal}(\x; \params_{\rm xtal}) &=
    \sum_{i=1}^{N_{\rm xtal}} A_{i, \rm xtal}
 \exp\left[-\frac{1}{2W_{i, \rm xtal}^2}\left(x - X_{i, \rm xtal}\right)^2\right].
\end{align}

\subsection{The amorphous component}
The amorphous component is also composed of gaussians, like the crystalline component.
In fact, the only real difference is the prior distributions for the positions,
amplitudes, and widths, given in Section~\ref{sec:priors}.

There is some number $N_{\rm amorph}$ of gaussians in the amorphous component.
The $i$th gaussian is parameterized
by its central position $X_{i, \rm amorph}$,
amplitude $A_{i, \rm amorph}$, and width $W_{i, \rm amorph}$.
The parameter vector for wide gaussians is
\begin{align}
\params_{\rm amorph} &=
  \{N_{\rm amorph}, X_1, X_2, ..., X_{N, \rm amorph},
    A_1, A_2, ..., A_{N, \rm amorph},
    W_1, W_2, ..., W_{N, \rm amorph}\}
\end{align}
and the total shape contributed by the
amorphous component is a sum of the $N_{\rm amorph}$ gaussians:
\begin{align}
f_{\rm amorph}(\x; \params_{\rm amorph}) &=
    \sum_{i=1}^{N_{\rm amorph}} A_{i, \rm amorph}
 \exp\left[-\frac{1}{2W_{i, \rm amorph}^2}\left(x - X_{i, \rm amorph}\right)^2\right].
\end{align}

@@

\begin{table}
\footnotesize
\centering
\begin{tabular}{|lll|}
\hline
{\bf Quantity}      &   {\bf Meaning}   &  {\bf Prior distribution}\\
\hline
{\em Prior information}&&\\
\hline
$n$ & Number of measurements & given\\
$\{\x_1, \x_2, ..., \x_n\}$  & $\x$-values of measurements & given \\
\hline
{\em Hyperparameters} & &\\
$N_{\rm xtal}$   &   Number of narrow peaks    &  $\propto 1/(N_{\rm xtal}+1)$, $N_{\rm xtal} \in \{0, 1, ..., 300\}$ \\
$N_{\rm amorph}$   &   Number of wide peaks    &  $\propto 1/(N_{\rm amorph}+1)$,
$N_{\rm amorph} \in \{0, 1, ..., 300\}$ \\
$a_{\ln A, \rm xtal}$ & Typical log-amplitude of narrow peaks & Laplace(0, 5)\\
$a_{\ln A, \rm amorph}$ & Typical log-amplitude of wide peaks & Laplace(0, 5)\\
$b_{\ln A, \rm xtal}$ & Diversity of log-amplitude of narrow peaks & Uniform(0, 5)\\
$b_{\ln A, \rm amorph}$ & Diversity of log-amplitude of wide peaks & Uniform(0, 5)\\
$a_{\ln W, \rm xtal}$ & Typical log-width of narrow peaks & LogUniform$(10^{-3}x_{\rm range}, 0.05x_{\rm range})$\\
$a_{\ln W, \rm amorph}$ & Typical log-width of wide peaks & LogUniform$(0.05x_{\rm range}, x_{\rm range})$\\
$b_{\ln W, \rm xtal}$ & Diversity of log-width of narrow peaks & Uniform(0, 0.1)\\
$b_{\ln W, \rm amorph}$ & Diversity of log-width of wide peaks & Uniform(0, 0.2)\\
\hline
{\em Background parameters}&&\\
$b$       & Mean background level       & $\ln b \sim $ Laplace$(0, 5)$\\
$\{n_1, ..., n_4\}$  & Background deviations & independent Normal$(0,1)$\\
\hline
{\em Crystalline parameters}&&\\
$X_{i, \rm xtal}$ & Positions of narrow peaks &
                            Uniform$(x_{\rm min}, x_{\rm max})$ \\
$A_{i, \rm xtal}$ & Amplitudes of narrow peaks &
 $\ln A_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln A, \rm xtal}, b_{\ln A, \rm xtal}\right)$ \\
$W_{i, \rm xtal}$ & Widths of narrow peaks &
 $\ln W_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln W, \rm xtal}, b_{\ln W, \rm xtal}\right)$ \\
\hline
{\em Amorphous parameters}&&\\
$X_{i, \rm amorph}$ & Positions of wide peaks &
          Uniform$(x_{\rm min}+0.3x_{\rm range}, x_{\rm max}-0.3x_{\rm range})$ \\
$A_{i, \rm amorph}$ & Amplitudes of wide peaks &
 $\ln A_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln A, \rm amorph}, b_{\ln A, \rm amorph}\right)$ \\
$W_{i, \rm amorph}$ & Widths of wide peaks &
 $\ln W_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln W, \rm amorph}, b_{\ln W, \rm amorph}\right)$ \\
\hline
{\em Noise parameters}&&\\
$\sigma_0$ &    Constant noise level  &   $\ln \sigma_0 \sim \textnormal{Laplace}(0,5)$\\
$\sigma_1$ &    Noise proportional to $\sqrt{\textnormal{model}}$   &  $\ln \sigma_1 \sim \textnormal{Laplace}(0,5)$ \\
$\nu$     &   Shape parameter for $t$-distribution for noise   &   LogUniform$(1, 1000)$\\
\hline
{\em Data}&&\\
\hline
$\{y_1, y_2, ..., y_n\}$  &   Measurements    & independent Student-$t(f(x_i; \params), \sqrt{\sigma_0^2 + \sigma_1f(x_i; \params)},\nu)$\\
\hline
\end{tabular}
\caption{\it The prior distributions for all hyperparameters,
parameters, and the data.\label{tab:priors}}
\end{table}

@@

\section{Applications}\label{sec:applications}
\subsection{Crystallinity}
The AMORPH program calculates a fraction of amorphous material ($A_{\rm m}$) for each posterior sample of the parameters,
using Equation~\ref{eqn:crystallinity}. Thus, measured crystallinity ($C_{\rm m}$) is easily determined from the relationship $C_{\rm m} = 1-A_{\rm m}$, as defined in Equation 5. The measured crystallinity is not equivalent to an ``actual'' or ``true'' crystallinity, although the values are often very close to one another.
It should therefore be noted that this procedure does not negate the need for calibration standards
to translate measured crystallinity to an actual crystallinity. Transmission and absorption of X-rays in the material will vary as the ratio of mineral to glass changes in the sample, causing a deviation from a 1:1 line. In addition to theoretical considerations, practical challenges also exist for background fitting. At low crystallinity ($\lesssim 10\%$) diffraction patterns are dominated by the low amplitude amorphous peak such that removing the effects of background noise becomes problematic, resulting in overestimated crystallinities. At high crystallinity ($\gtrsim 90\%$) the problem arises that peak overlaps in multi-mineral assemblages cause broadening effects, and may be subsequently modeled as ``amorphous'' and thus crystallinities  underestimated. The calibration curve therefore is utilized to remove these effects. This is also an important step as
different X-ray diffractometers will provide slightly different calibration curves (Fig.~\ref{fig:intruments}).
Differences between instruments largely reflect changes in background intensity and shape,
particularly at low values of $2\theta$.

>>>>

\documentclass[review]{elsarticle}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{dsfont}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
% \usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{pdflscape}
\usepackage{gensymb} % For degree symbol
\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\journal{Computers and Geosciences}

@@

\begin{abstract}
AMORPH utilizes a new Bayesian statistical approach to interpreting X-ray diffraction results of samples with both crystalline and amorphous components. AMORPH fits X-ray diffraction patterns with a mixture of gaussians,
which allows us to simultaneously infer all of the model parameters and quantify their
uncertainties.
The program simulates background patterns previously manually applied, providing reproducible results, and significantly reducing inter- and intra-user biases. This approach allows for the quantification of amorphous and crystalline materials and for the characterization of the amorphous component, including parameters of centre of mass, width, skewness, and nongaussianity of the amorphous component. Results demonstrate the applicability of this program for calculating amorphous contents of volcanic materials and independently modeling their properties in compositionally variable materials.
\end{abstract}

@@

\linenumbers

\section{Introduction}
Quantification and characterization of amorphous materials is an important area of research in
terrestrial and planetary sciences \citep[e.g.,][]{schmidt2009, dehouck2014, wall2014, morris2016, zorn2018}.
Powder X-ray diffraction (XRD), although primarily a tool for characterization of crystalline materials,
also provides a means to investigate amorphous materials \citep[e.g.,][]{rowe2012, achilles2013}. X-ray amorphous materials lack long-range crystallographic order and produce broad humps with low intensities in XRD patterns. X-ray diffraction of rocks/soils containing amorphous materials (e.g. volcanic
glass, amorphous silica, and organics) produces a diffraction pattern which is a combination of the
crystalline and amorphous phases. Recent studies have debated the nature of amorphous materials in
planetary sciences \citep[e.g.][]{schmidt2009, dehouck2014}, and a growing number of
applications to terrestrial volcanics have presented an avenue of expansion for XRD methodologies
for amorphous material analysis \citep{ellis2015, kanakiya2017, zorn2018}.
Here we present a new statistical approach and computer program, AMORPH, which utilizes Bayesian inference to determine
the crystallinity, and other characteristics, of partially amorphous materials from X-ray diffraction patterns, as well as the uncertainty in these quantities.

Several techniques currently exist for the quantification of amorphous materials in heterogeneous (amorphous + crystalline) materials. However, these
each present their own limitations. The two most commonly used approaches in recent literature
either rely on a calibration curve for different crystallinities \citep[calibration method;][]{ rowe2012} or a combined Rietveld-Reference Intensity Ratio (RIR) method \citep{gualtieri1996, gualtieri2000}. While the combined Rietveld-RIR procedure allows for quantification of both the amorphous 
and crystalline materials, for proper characterization it requires a library of pure components,
spiked with a reference material (e.g. corundum). Sample spiking presents additional complications
as it reduces the observed intensity of the amorphous material thus reducing the accuracy of
pattern matching to determine abundances. An Excel based adaptation of this pattern matching
technique \citep[FULLPAT;][]{chipera2002} is routinely utilized for characterizing remotely acquired X-ray
diffraction patterns, such as those measured by the CheMin instrument on the Mars Science Laboratory (MSL) Curiosity rover \citep{blake2012, bish2013}. The problem faced by RIR approaches is they are based on modeling of an observed 
diffraction pattern compared to library measurements, and therefore do not allow an independent
assessment of the characteristics of the amorphous material. Furthermore, natural X-ray amorphous phases have a wide variety of compositions such that the library measurements may not include the amorphous phase(s) present in the sample. In comparison, the calibration method
focuses solely on the quantification of the amorphous material, utilizing the integrated counts
associated with the amorphous and crystalline componentry.

@@

\section{Bayesian inference}\label{sec:bayes}
AMORPH is founded upon Bayesian inference
\citep{gregory2005bayesian, o2004kendall, sivia2006data}, where
probability distributions are used to model
states of uncertainty about the values of unknown quantities.
Bayesian inference is often implemented computationally with
Markov Chain Monte Carlo (MCMC) methods \citep{mackay2003}, and AMORPH is no exception.
Loosely speaking, the goal is to fit a dataset with a model by exploring
those regions of the model's parameter space that are consistent with the
data. The model assumptions are described in detail in
Section~\ref{sec:model}, and the computation is performed with
Diffusive Nested Sampling, an advanced MCMC method \citep{dns, dnest4}.
There have been many applications of Bayesian inference to
problems of separating background from compact signals
when noise is present \citep[e.g.][]{}. AMORPH can be viewed as another
case of this, with assumptions customized to the specific application.

@@

The prior describes an initial state of uncertainty about the
parameter values, and is often a wide probability distribution.
The prior is supplemented with $p(\data | \params, \info)$, sometimes
called a `sampling distribution',
which describes the uncertainty about the data which we {\em would} have,
if we knew the parameters (and the prior information), as a function of
the parameters. This describes uncertainty about what data will be observed,
but encodes some knowledge about the fact that the data have something to do with the
parameters $\params$. For example, $p(\data | \params, \info)$ often encodes
the assumption that the data has noise in it, and the specific noise values
are unknown a priori.

@@

When $\params$ is a large collection of parameters, the posterior distribution
is a probability distribution over a high-dimensional parameter space. To
represent it computationally, Markov Chain Monte Carlo (MCMC) methods are
often used to generate samples from the parameter space, according to the
posterior distribution. The output is a collection of plausible values of the
parameters which can be used to approximate posterior probabilities.
For example, the probability some parameter $\phi$ is greater than 3.4
is approximately the proportion of the posterior samples that have
$\phi > 3.4$. While many data analysis procedures work by finding the
best fitting values of the parameters (i.e., a maximum likelihood estimate),
Bayesian inference via MCMC involves finding a sampling of plausible values
for the parameters. Since AMORPH uses \citet{dnest4}, it also calculates
the `marginal likelihood' or `evidence' quantity $p(\data | \info)$, which
quantifies how well the model fits the data overall, averaged over all
possible values of its parameters. This allows straightforward model comparison
with any alternatives that may be proposed in the future
\citep{skilling2006nested}.

% BJB: Hopefully this addresses reviewer 1, point a.
The computational cost of running AMORPH on a data set of around 1000
points varies, as the posterior distribution depends on the data.
Another important factor is the number of MCMC samples desired.
On a modern PC, some useful posterior samples can be obtained in about
half an hour. With 300 posterior samples, typical analyses take approximately 7 hours to complete, depending on computing capabilities and the number of data points. However, the complete runs presented in this paper took
about a day each to run, as we wanted to ensure the results contained
about 1000 posterior samples.

@@

\subsection{The background component}
The background component is assumed to be piecewise linear, with constant
slope
{between $\x=\x_{\rm min}$ and $\x=10$\degree, between $\x=10$\degree~and $\x=40$\degree,
and between $\x=40$\degree~and $\x=\x_{\rm max}$}. These values are selected to maintain a linear background over the interval from 10--40\degree, the predominant range of amorphous materials (8.8 -- 2.25 \AA) analyzed using a Cu K$\alpha$ x-ray source (Table~\ref{tab:anode}), while still allowing for some flexibility in the structure of the background outside this range.

We parameterized the background component using
five parameters. Firstly, there is a mean level $b$, which sets the
typical amplitude of the background component. There are also four parameters
$n_1, ..., n_4$, which determine the offset of four `control points'
from the mean level. The positions of the four control points are then:
\begin{align}
\left(\x_{\rm min}, be^{n_1}\right) \nonumber\\
\left(10, be^{n_2}\right) \nonumber\\
\left(40, be^{n_3}\right) \nonumber\\
\left(\x_{\rm max}, be^{n_4}\right). \nonumber
\end{align}
Control points may be modified, depending on the X-ray source in use and changes in the required range for the linear background as a function of the nature of the amorphous material (Table~\ref{tab:anode}). The value of the background component at any other point is found by
linear interpolation of these control points.
Figure~\ref{fig:background} shows three example
background curves generated from this parameterization, with typical
values of the parameters.

\begin{table}
\footnotesize
\centering
\begin{tabular}{|llll|}
\hline
{\bf X-ray Anode}      &   {\bf X-ray wavelength}   &  {\bf Control point 2} & {\bf Control point 3}\\
 &   {\bf (\AA)}   &  {\bf(\degree 2 theta)} & {\bf(\degree 2 theta)}\\
\hline
Cr & 2.29 & 15.0 & 61.2\\
Fe & 1.94 & 12.7 & 51.1\\
Co & 1.79 & 11.7 & 46.9\\
{\bf Cu} & {\bf 1.54} & {\bf 10.0} & {\bf 40.0}\\
Mo & 0.71 & 4.6 & 18.2\\
Ag & 0.56 & 3.6 & 14.3\\
\hline
\end{tabular}
\caption{\it Typical XRD anodes with recommended background control points based on a predominant 8.8-2.25 {\AA}  amorphous peak.\label{tab:anode}}
\end{table}

@@

\subsection{The crystalline component}

% BJB: Reponse to Reviewer 2, comment about Lorentzian shape vs Gaussian
% We experimented with the Lorentzian shape instead of the Gaussian and found
% that for same datasets, it was preferred, whereas for other datasets, the
% Gaussian is preferred (as adjudicated by the marginal likelihood values
% for the two different assumptions). Rather than having to run each model
% twice, we decided it would be more elegant to replace the gaussian assumption
% with some family of shapes that can smoothly interpolate between gaussian
% and lorentzian as necessary. Specifically, we chose the Student's
% t-distribution which can do this straightforwardly. This introduces one
% more parameter to the model, which controls the shape of the peaks.

% I (BJB) also changed the parameterisation of the heights of the peaks, so
% they are described by the total integrated mass rather than the peak
% amplitude. This makes the Metropolis exploration a bit more efficient.

We assume that there is some number $N_{\rm xtal}$ of narrow peaks
due to crystalline material,
each of which has the same shape. The shapes of the narrow
peaks are controlled by a single shape parameter $\nu_{\rm xtal}$,
assumed common to all of the narrow peaks. The chosen shape is based
on the Student's $t$-distribution, which allows to peaks to be heavy-tailed
if $\nu_{\rm xtal}$ is low, and to be Lorentzian or Cauchy in shape if
$\nu_{\rm xtal}$ is exactly 1. As $\nu_{\rm xtal}$ is high, the shape
approximates a gaussian.

The $i$th peak is parameterized
by its central position $X_{i, \rm xtal}$,
total integrated `mass' $M_{i, \rm xtal}$, and width $W_{i, \rm xtal}$.
The parameter vector for crystalline component is
\begin{align}
\params_{\rm xtal} &=
  \{N_{\rm xtal}, \nu_{\rm xtal}, X_{1, \rm xtal}, ..., X_{N, \rm xtal},
    M_{1, \rm xtal}, ..., M_{N, \rm xtal},
    W_{1, \rm xtal}, ..., W_{N, \rm xtal}\}
\end{align}
and the total shape contributed by the
crystalline component is a sum of the $N_{\rm xtal}$ peaks:
\begin{align}
f_{\rm xtal}(\x; \params_{\rm xtal}) &=
	\frac{\Gamma(\frac{\nu_{\rm xtal} + 1}{2})}{\sqrt{\nu_{\rm xtal}\pi}
                 		\Gamma(\frac{\nu_{\rm xtal}}{2})}
    \sum_{i=1}^{N_{\rm xtal}}
    		\frac{M_{i, \rm xtal}}
            	 {W_{i, \rm xtal}}
	\left(1 + \frac{1}{\nu}\left(\frac{x - X_{i, \rm xtal}}{W_{i, \rm xtal}}\right)^2\right)
    			^{-\frac{\nu_{\rm xtal} + 1}{2}}. \label{eqn:t_shape}
\end{align}
The terms inside the sum have the form of the probability density function
of the $t$-distribution, adjusted by shifting according to the central
position and widening by the width, and the total integral is scaled by
the mass parameters. $\Gamma()$ is the gamma function.

For computational efficiency, evaluation of Equation~\ref{eqn:t_shape}
is performed using a lookup table.

\subsection{The amorphous component}
The amorphous component is also composed of a sum of peaks, which we chose to be
gaussians. The parameterization is the same as for the crystalline peaks but with a fixed,
high value for the shape parameter.
Apart from the gaussian shape, the only other difference is the prior distributions for the positions, masses, and widths, given in Section~\ref{sec:priors}. In short, the wide gaussians representing the
amorphous component are more likely to be located near the middle of the
dataset, and to be wider.

There is some number $N_{\rm amorph}$ of gaussians in the amorphous component.
The $i$th gaussian is parameterized
by its central position $X_{i, \rm amorph}$,
mass (total integral) $M_{i, \rm amorph}$, and width $W_{i, \rm amorph}$.
The parameter vector for wide gaussians is
\begin{align}
\params_{\rm amorph} &=
  \{N_{\rm amorph}, X_{1, \rm amorph}, ..., X_{N, \rm amorph},
    M_{1, \rm amorph}, ..., M_{N, \rm amorph},
    W_{1, \rm amorph}, ..., W_{N, \rm amorph}\}
\end{align}
and the total shape contributed by the
amorphous component is a sum of the $N_{\rm amorph}$ gaussians:
\begin{align}
f_{\rm amorph}(\x; \params_{\rm amorph}) &=
	\frac{1}{\sqrt{2\pi}}
    \sum_{i=1}^{N_{\rm amorph}} \frac{M_{i, \rm amorph}}{W_{i, \rm amorph}}
 \exp\left[-\frac{1}{2W_{i, \rm amorph}^2}\left(x - X_{i, \rm amorph}\right)^2\right].
\end{align}

@@

\begin{table}
\footnotesize
\centering
\begin{tabular}{|lll|}
\hline
{\bf Quantity}      &   {\bf Meaning}   &  {\bf Prior distribution}\\
\hline
{\em Prior information}&&\\
\hline
$n$ & Number of measurements & given\\
$\{\x_1, \x_2, ..., \x_n\}$  & $\x$-values of measurements & given \\
\hline
{\em Hyperparameters} & &\\
$N_{\rm xtal}$   &   Number of narrow peaks    &  $\propto 1/(N_{\rm xtal}+1)$, $N_{\rm xtal} \in \{0, 1, ..., 1000\}$ \\
$N_{\rm amorph}$   &   Number of wide peaks    &  $\propto 1/(N_{\rm amorph}+1)$,
$N_{\rm amorph} \in \{0, 1, ..., 1000\}$ \\
$a_{\ln M, \rm xtal}$ & Typical log-mass of narrow peaks & Laplace(0, 5)\\
$a_{\ln M, \rm amorph}$ & Typical log-mass of wide peaks & Laplace(0, 5)\\
$b_{\ln M, \rm xtal}$ & Diversity of log-mass of narrow peaks & Uniform(0, 5)\\
$b_{\ln M, \rm amorph}$ & Diversity of log-mass of wide peaks & Uniform(0, 5)\\
$a_{\ln W, \rm xtal}$ & Typical log-width of narrow peaks & LogUniform$(10^{-3}x_{\rm range}, 0.05x_{\rm range})$\\
$a_{\ln W, \rm amorph}$ & Typical log-width of wide peaks & LogUniform$(0.05x_{\rm range}, x_{\rm range})$\\
$b_{\ln W, \rm xtal}$ & Diversity of log-width of narrow peaks & Uniform(0, 0.1)\\
$b_{\ln W, \rm amorph}$ & Diversity of log-width of wide peaks & Uniform(0, 0.2)\\
\hline
{\em Background parameters}&&\\
$b$       & Mean background level       & $\ln b \sim $ Laplace$(0, 5)$\\
$\{n_1, ..., n_4\}$  & Background deviations & independent Normal$(0,1)$\\
\hline
{\em Crystalline parameters}&&\\
$X_{i, \rm xtal}$ & Positions of narrow peaks &
                            Uniform$(x_{\rm min}, x_{\rm max})$ \\
$A_{i, \rm xtal}$ & Masses of narrow peaks &
 $\ln M_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln M, \rm xtal}, b_{\ln M, \rm xtal}\right)$ \\
$W_{i, \rm xtal}$ & Widths of narrow peaks &
 $\ln W_{i, \rm xtal} \sim \textnormal{Laplace}\left(a_{\ln W, \rm xtal}, b_{\ln W, \rm xtal}\right)$ \\
$\nu_{\rm xtal}$ & Shape of narrow peaks (common) & LogUniform$(0.1, 100)$\\
\hline
{\em Amorphous parameters}&&\\
$X_{i, \rm amorph}$ & Positions of wide peaks &
          Uniform$(x_{\rm min}+0.3x_{\rm range}, x_{\rm max}-0.3x_{\rm range})$ \\
$A_{i, \rm amorph}$ & Masses of wide peaks &
 $\ln M_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln M, \rm amorph}, b_{\ln M, \rm amorph}\right)$ \\
$W_{i, \rm amorph}$ & Widths of wide peaks &
 $\ln W_{i, \rm amorph} \sim \textnormal{Laplace}\left(a_{\ln W, \rm amorph}, b_{\ln W, \rm amorph}\right)$ \\
\hline
{\em Noise parameters}&&\\
$\sigma_0$ &    Constant noise level  &   $\ln \sigma_0 \sim \textnormal{Laplace}(0,5)$\\
$\sigma_1$ &    Noise proportional to $\sqrt{\textnormal{model}}$   &  $\ln \sigma_1 \sim \textnormal{Laplace}(0,5)$ \\
$\nu$     &   Shape parameter for $t$-distribution for noise   &   LogUniform$(1, 1000)$\\
\hline
{\em Data}&&\\
\hline
$\{y_1, y_2, ..., y_n\}$  &   Measurements    & independent Student-$t(f(x_i; \params), \sqrt{\sigma_0^2 + \sigma_1f(x_i; \params)},\nu)$\\
\hline
\end{tabular}
\caption{\it The prior distributions for all hyperparameters,
parameters, and the data.\label{tab:priors}}
\end{table}

@@

\section{Applications}\label{sec:applications}
\subsection{Crystallinity}
To date, applications of crystallinity determinations via XRD methodologies on terrestrial samples have primarily focused on volcanic materials \citep[e.g.,][]{wall2014, ellis2015, andrade2017, zorn2018}, linking crystallinity of volcanic glass to timescales of geologic processes. The primary advantage of this technique being its ability to determine crystallinities for finer grain sizes than typically possible through standard petrologic techniques \citep{rowe2012}. In contrast, planetary science usages for determining crystallinity have largely focused on quantification of the amorphous and crystalline components for mass balance determination of physical and chemical components in diverse geologic materials \citep[e.g.,][]{blake2013, dehouck2014}.

The AMORPH program calculates a fraction of amorphous material ($A_{\rm m}$) for each posterior sample of the parameters,
using Equation~\ref{eqn:crystallinity}.Thus measured crystallinity ($C_{\rm m}$) is easily determined from the relationship $C_{\rm m}$ = 1-$A_{\rm m}$, as defined in Equation 5. 
To test the applicability of AMORPH crystallinity results we have analysed the Mars Scientific Laboratory (MSL)  CheMin instrument X-ray diffraction patterns of loose basaltic sand from the Bagnold Sand Dunes \citep[Gobabeb;][]{achilles2017, lapotre2017}
and sediments drilled from a silicic mudstone \citep[Buckskin sample;][]{morris2016}, both which have significant amorphous
materials (CheMin instrument data\footnote{\tt http://pds-geosciences.wustl.edu/msl/msl-m-chemin-4-rdr-v1/mslcmn{\textunderscore}1xxx.data.rdr4/}). CheMin diffraction patterns have been processed using the original Co K$\alpha$ ($1.7890 \dot{A}$) data as well as data transformed from a Co K$\alpha$ to Cu K$\alpha$ ($1.5406 \dot{A}$) X-ray source using Bragg's law. This recalculation simply allows for an easier visual comparison to terrestrial samples analysed with the more standard Cu K$\alpha$ X-ray source (see below). Note that for Co K$\alpha$ results, background critical points have been adjusted to 11.7-46.9 $2\theta$ to maintain the linear background over the amorphous region from ~8.8 to 2.25 $\dot{A}$ (Table ~\ref{tab:anode}). Although the results are uncalibrated, measured crystallinites provide a good first approximation of actual crystallinity, and relative differences are robust. Measured amorphous content of the Gobabeb 2 sample is 36.7 $\pm$ 2.9$\%$ (63.3$\%$ crystallinity) while the Buckskin 2 sediment contains 45.3 $\pm$ 2.9$\%$ amorphous phases (54.7 $\%$ crystallinity; Fig.~\ref{fig:Mars}). Comparatively, published FULLPAT calculations of amorphous contents are ~35 $\pm$ 15$\%$ for Gobabeb and 50 $\pm$ 15$\%$ for Buckskin \citep{achilles2017, morris2016}.

\begin{figure}[!ht]
\centering
\includegraphics[width=1\textwidth]{figures/Mars.jpg}
\caption{\it Analysis of Gobabeb 2 (left) and Buckskin 2 (right) results from the CheMin instrument on the MSL Curiosity rover (data sourced as above). Upper panels  illustrate initial Co K$\alpha$ processed results while lower panels show AMORPH results after translating data to a Cu K$\alpha$ source (see text for details). Calculated statistics provided by AMORPH for each analysis are indicated below the model results.\label{fig:Mars}}
\end{figure}
